---
title: "CogLab Data Pipeline"
subtitle: "From raw files to interpretable visualizations"
author: "Ryan Mears and LLM"
institute: "University of Florida"
format:
  revealjs:
    theme:
      - default
      - garfield.scss
    slide-number: true
    incremental: true
    transition: fade
    transition-speed: slow
    navigation-mode: vertical
    controls-layout: bottom-right
    controls-tutorial: true
    center: false
    smaller: true

execute:
  echo: true
  warning: false
  message: false
---

 

```{r}
#| echo: false


library(tidyverse)
library(janitor)

garfield_cols <- c(
  orange = "#F28C28",
  dark_orange = "#D96A1B",
  cream = "#F6E7C1",
  beige = "#FAF2DA",
  brown = "#3A2A1A",
  tan = "#E8D3A7",
  red = "#B7410E"
)

scale_color_garfield <- function(...) {
  ggplot2::scale_color_manual(
    values = c("#F28C28", "#D96A1B", "#B7410E", "#A66A2C", "#6E4B2A"),
    ...
  )
}

scale_fill_garfield <- function(...) {
  ggplot2::scale_fill_manual(
    values = c("#F28C28", "#D96A1B", "#E8D3A7", "#F6E7C1", "#B7410E"),
    ...
  )
}

theme_garfield <- function(base_size = 18, base_family = "sans") {
  ggplot2::theme_minimal(base_size = base_size, base_family = base_family) +
    ggplot2::theme(
      plot.background = ggplot2::element_rect(fill = "#FAF2DA", color = NA),
      panel.background = ggplot2::element_rect(fill = "#F6E7C1", color = NA),
      panel.grid.major = ggplot2::element_line(color = "#E8D3A7", linewidth = 0.5),
      panel.grid.minor = ggplot2::element_blank(),
      plot.title = ggplot2::element_text(face = "bold", color = "#B7410E", size = base_size * 1.2),
      plot.subtitle = ggplot2::element_text(color = "#6E4B2A"),
      axis.title = ggplot2::element_text(color = "#3A2A1A", face = "bold"),
      axis.text = ggplot2::element_text(color = "#3A2A1A"),
      legend.title = ggplot2::element_text(face = "bold", color = "#3A2A1A"),
      legend.text = ggplot2::element_text(color = "#3A2A1A"),
      strip.background = ggplot2::element_rect(fill = "#F28C28", color = "#D96A1B"),
      strip.text = ggplot2::element_text(color = "#3A2A1A", face = "bold")
    )
}

```

# Overview


::: {.bubble-title}

Deck purpose

:::

This presentation shows a practical sequence for moving from imported CogLab data to <br /> tidy variables, validated summaries, and clearer interpretation.



::: notes

Introduce the deck as a workflow map rather than just a sequence of code steps.
Emphasize that students often underestimate the importance of cleaning and validation.

:::

# Overview

::: {.speech-bubble}
::: {.bubble-title}

Why this workflow matters

:::

Raw CogLab data usually do **not** begin in a form ready for interpretation.

We need a pipeline to move from **files** to **tidy variables** to **validated summaries**.

:::

::: {.pipeline .center} 

<span class="pipeline-step active fragment">Provision</span> <span class="pipeline-arrow fragment">➡</span> <span class="pipeline-step fragment">Import</span> <span class="pipeline-arrow fragment">➡</span> <span class="pipeline-step fragment">Tidy</span> <span class="pipeline-arrow fragment">➡</span> <span class="pipeline-step fragment">Transform</span> <span class="pipeline-arrow fragment">➡</span> <span class="pipeline-step fragment">Visualize/Validate</span> <span class="pipeline-arrow fragment">➡</span> <span class="pipeline-step fragment">Re-Visualize</span>

:::

::: notes
Frame the pipeline as a sequence of decisions.
You can mention that most effort in real projects occurs before inferential statistics.
:::

# Provision packages

## Slide 1

::: {.pipeline .center} 

<span class="pipeline-step active">Provision</span> <span class="pipeline-arrow">➡</span> <span class="pipeline-step">Import</span> <span class="pipeline-arrow">➡</span> <span class="pipeline-step">Tidy</span> <span class="pipeline-arrow">➡</span> <span class="pipeline-step">Transform</span> <span class="pipeline-arrow">➡</span> <span class="pipeline-step">Visualize/Validate</span> <span class="pipeline-arrow">➡</span> <span class="pipeline-step">Re-Visualize</span>

:::

::: {.speech-bubble}
::: {.bubble-title}

Goal

:::

Load packages needed for wrangling, checking, and plotting.

:::

```{r}
library(tidyverse)
library(janitor)
library(readr)
library(stringr)
library(forcats)
```

::: notes
Explain why package provisioning is part of reproducibility.
Students should learn that dependencies are part of the method.
:::

# Import data

## Slide 2

::: {.pipeline .center} 

<span class="pipeline-step">Provision</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step active">Import</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Tidy</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Transform</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Visualize/Validate</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Re-Visualize</span>

:::


::: {.speech-bubble}
::: {.bubble-title}

Import question

:::

What files exist, and what variables arrive from CogLab in raw form?

:::

```{r}
data_raw <- tibble(
  subj = c("s01", "s02", "s03", "s04"),
  Cond = c("Valid", "Invalid", "Valid", "Invalid"),
  RT_ms = c(510, 640, 490, 710),
  Accuracy = c(1, 0, 1, 1)
)

data_raw
```

::: notes

Point out that imported names and types are often inconsistent with analysis needs.
This is a good place to discuss file formats and provenance.

:::

# Tidy names and data types

## Slide 3

::: {.pipeline .center} 

<span class="pipeline-step">Provision</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Import</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step active">Tidy</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Transform</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Visualize/Validate</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Re-Visualize</span>

:::

::: {.speech-bubble}
::: {.bubble-title}

Cleaning question

:::

Are names consistent and are variable classes appropriate for analysis?

:::

```{r}
data_tidy <- data_raw |>
  janitor::clean_names() |>
  mutate(
    cond = factor(str_to_lower(cond)),
    accuracy = as.logical(accuracy)
  )

glimpse(data_tidy)
```

::: notes

Explain why tidy naming reduces downstream confusion.
Mention that factor, logical, numeric, and character types imply different kinds of operations.

:::

# Transform

## Slide 4

::: {.pipeline .center} 

<span class="pipeline-step">Provision</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Import</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Tidy</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step active">Transform</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Visualize/Validate</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Re-Visualize</span>

:::

::: {.speech-bubble}
::: {.bubble-title}

Transform question

:::

What derived variables help us understand performance?

:::

```{r}
data_trans <- data_tidy |>
  mutate(
    rt_s = rt_ms / 1000,
    acc_num = as.integer(accuracy)
  )

data_trans
```

::: notes

Use this slide to distinguish direct measurements from derived variables.
This is often where methods sections need extra specificity.

:::

# Visualize and Validate

## Slide 5

::: {.pipeline .center} 

<span class="pipeline-step">Provision</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Import</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Tidy</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Transform</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step active">Visualize/Validate</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Re-Visualize</span>

:::



::: {.speech-bubble}
::: {.bubble-title}

First look

:::

Before modeling, inspect distributions and obvious condition differences.

:::


```{r}
#| echo: true
#| output-location: column-fragment

ggplot(data_trans, aes(x = cond, y = rt_ms, fill = cond)) +
  geom_boxplot(alpha = 0.9, width = 0.6) +
  scale_fill_garfield() +
  labs(
    title = "Response time by condition",
    x = "Condition",
    y = "RT (ms)"
  ) +
  theme_garfield(18)
```



::: notes

Stress that visualization is part of validation, not just presentation.
Students should see the plot as a diagnostic tool.

:::

## Slide 6

::: {.speech-bubble}
::: {.bubble-title}

Validation question

:::

Do the values look plausible, or are there impossible or suspicious observations?

:::

```{r}

data_trans |>
  summarise(
    n = n(),
    min_rt = min(rt_ms, na.rm = TRUE),
    max_rt = max(rt_ms, na.rm = TRUE),
    mean_rt = mean(rt_ms, na.rm = TRUE),
    accuracy_rate = mean(acc_num, na.rm = TRUE)
  )
```

::: notes

Summaries can reveal range problems that are not obvious from a plot.
Connect this to conditional validation rules.

:::

## Slide 7

::: {.speech-bubble}
::: {.bubble-title}

Flagging rule example

:::

A common rule is to identify implausibly short or long reaction times.

:::

```{r}
data_checked <- data_trans |>
  mutate(
    rt_flag = case_when(
      rt_ms < 150 ~ "too_fast",
      rt_ms > 2000 ~ "too_slow",
      TRUE ~ "ok"
    )
  )

data_checked
```

::: notes

Explain that thresholds should be justified and reported.
This is where arbitrary cleaning choices can become hidden assumptions.

:::

# Re-Visualize

## Slide 8

::: {.pipeline .center} 

<span class="pipeline-step">Provision</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Import</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Tidy</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Transform</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step">Visualize/Validate</span> <span class="pipeline-arrow">→</span> <span class="pipeline-step active">Re-Visualize</span>

:::

::: {.speech-bubble}
::: {.bubble-title}

After validation

:::

Plot again after defining checks so that interpretation reflects conditioned data.

:::

```{r}
#| fig-width: 20
#| fig-height: 10
#| out-width: 100%
#| fig-align: center
#| output-location: slide

ggplot(data_checked, aes(x = rt_ms, fill = rt_flag)) +
  geom_histogram(binwidth = 50, color = "#3A2A1A") +
  scale_fill_garfield() +
  labs(
    title = "RT distribution after validation rules",
    x = "RT (ms)",
    y = "Count",
    fill = "Flag"
  ) +
  theme_garfield(28)
```

::: notes

This is a good point to compare before and after.
Emphasize that preprocessing changes what later conclusions mean.

:::

# Discuss

## Slide 9

::: {.speech-bubble}
::: {.bubble-title}

Discussion prompts

:::

* Which data-cleaning decisions were **required** versus **optional**?
* Which transformations changed interpretation the most?
* What assumptions should be reported in the methods section?

:::

::: notes

Invite students to treat the pipeline as an argument, not just a recipe.
Each step should be explainable and defendable.

:::

## Slide 10

::: {.speech-bubble}
::: {.bubble-title}

Take-home message

:::

The analysis is only as interpretable as the pipeline that produced it.
Good visualization helps reveal whether the data are ready for inference.

:::

::: notes

Close by linking the pipeline to reproducibility, transparency, and methods reporting.

:::



## Credits 

- [Hero Fandom: Garfield](https://hero.fandom.com/wiki/Garfield)
